---
title: The Pleistocene-Holocene Transition in Tagua-Tagua, central Chile
author:
  - name: Matías Frugone-Álvarez
    email: matutefrugone@gmail.com
    footnote: 1
    affiliation: puc
  - name: Erwin Gonzáles Huerta
    affiliation: uccs,ciba,geup
  - name: Blas Valero-Garcés
    affiliation: ipe
  - name: Carolina Godoy
    affiliation: uccs,ciba,geup
  - name: Josu Aranbarri
    affiliation: uccs,ciba,geup
  - name: Penélope Gonzáles
    affiliation: iact
  - name: Antonio Delgado-Huerta
    affiliation: uccs,ciba,geup
  - name: Claudio Latorre
    affiliation: deuc
  - name: Rafael Labarca Encina
    affiliation: uccs,ciba,geup
address:
  - code: puc
    address: Pontificia Universidad Católica de Chile, Departamento de Ecología-Centro UC Desierto de Atacama, Av. Libertador Bernardo O'Higgins 340, Santiago, Chile.
  - code: ieb
    address: Instituto de Ecología y Biodiversidady (IEB), Santiago, Chile
  - code: uccs
    address: Instituto Pirenaico de Ecología (IPE-CSIC), Zaragoza, Spain.
  - code: ipe
    address: Centro de Investigación en Biodiversidad y Ambientes Sustentables (CIBAS), Chile 
  - code: geup
    address: Department of Geology and Environmental Science, University of Pittsburgh, Pittsburgh, USA      
  - code: iact
    address: Instituto Andaluz de Ciencias de la Tierra (CSIC-UGR), Granada, Spain  
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  Paleohydrological reconstructions from the Chilean Altiplano document abrupt moisture fluctuations during the last millennia. Although the end of the mid Holocene aridity and the onset of more humid conditions between 6--4 ka has been identified in numerous regional marine and terrestrial sites, the timing of late Holocene dry and humid phases shows large regional variability. Laguna Miscanti and Laguna Miñiques (23$^\circ$ 43'S -- 67$^\circ$ 46'W, 4140 m asl) are topographically closed, but connected by surface outflow from Miscanti. Sedimentological and geochemical indicators from two new cores show large facies changes, i.e. higher carbonate and evaporite deposition during more arid periods and increased organic productivity (both algal and macrophyte) during more humid phases. As in most Andean lakes located in volcanic settings, large 14C reservoir effects occur complicating 14C dating, so the age models include 210Pb and U/Th dating. In spite of dating uncertainties, both lakes show similar patterns during the last millennium. A humid phase in Laguna Miscanti prior to ca 1200 CE is coherent with rodent middens and geomorphological features indicative of a major pluvial/recharge event at lower altitudes (Atacama Desert) during the Medieval Climate Anomaly (ca 800 - 1300 CE). The LIA (1300 – 1850 CE) is characterized by several arid/humid cycles and the last century by a productivity increase. The hydrological changes observed during the last millennium illustrate the complex dynamics of recent climate evolution over the high altitude Andean plateau. Discrepancies between the northern and southern Altiplano records and with intermediate latitudes (Central Chile) records may reflect contrasting responses to external forcing (Westerlies versus South American Monsoon dynamics) along different climatic zones.
  
keywords: "keywords"
journal: "Quaternary Science Reviews"
date: "`r Sys.Date()`"
bibliography: mybibfile.bib
linenumbers: true
numbersections: true
csl: elsevier-harvard.csl
output: rticles::elsevier_article
header-includes:
- \usepackage{caption}
- \usepackage{multirow}
- \usepackage{tabularx}
---

```{r include = FALSE}

library(tidyverse)
library(targets)

```

```{r include = FALSE}

options(tinytex.verbose = TRUE)

```

# Introduction

# Study site and regional setting

# Materials and methods

In November 2019, we recovered a sedimentary sequence of 15 cm length,
10 cm width and 280 cm depth of the profile exposed on the western wall
of a 1x2 m excavation in the ALTT (Tagua-Tagua 3, Figure 1). Sediment
samples (5 to 10 g of material) were obtained at 1-cm intervals.
Sixty-six samples were sent to the laboratories of Pyrenees Ecology
Institute (IPE-CSIC) and the Andalusian Institute Earth Sciences
Institute (IACT-CSIC) in Spain for geochemical, pollen and isotopic
analysis. Another set of samples were sent to the Archaeobotany and
Palaeoecology Laboratory of the University of Exeter for the processing
and extraction of diatom and phytolith. Sedimentary facies were defined
and characterized based on macroscopic description (color, bedding
features, grain size, lithology and sedimentary textures) and a machine
learning workflow using the scikit-learn library (cita) in Python.
Sedimentary facies were defined using a dimensionality reduction
(Exploratory Factor Analysis) and unsupervised machine learning (K-Means
algorithm cluster) with three different dataset (elemental geochemistry,
grain size and diatoms)(Hall., 2016).

Samples were taken every 2 cm for analysed of total carbon (TC), total
inorganic carbon (TIC), total organic carbon ($TOC = T-TIC$), total
nitrogen (TN) and total sulfur (TS) in a LECO SCN furnace (IPE--CSIC).
The grain-size frequency distributions (between 0.02 um and 2000 um)
were measured in triplicate using a Malvern Mastersizer 2000 APA2000
grain-size analyzer with Hydro 2000G AWA2000. To determine the
grain--size, the samples were pre-treated with 10% H2O2 and 10% HCl to
remove organic matter and carbonates, respectively. Robust End-member
mixing analysis (EMMA) was performed using the generic deterministic
EMMA protocol with the R package EMMAgeo v0.9.8. (Dietze et al 2012;
Dietze and Dietze 2019) to infer the sediment sources, transport and
depositional processes. Stable carbon and oxygen isotope analyses on
calcite were carried out every 2 cm by mass spectrometry in bulk
sediment samples following standard procedures and using an IRMS
Finnigan MAT 251. The isotopic values are reported in the conventional
delta notation relative to the PDB standard.

Eleven pollen samples were processed following Faegri and Iversen (1989)
methods to extract pollen grains. Lycopodium tablets were added for
calculating pollen concentration (grains/cm3; Stockmarr(1971)) and
accumulation rates (grains/cm2/yr). Pollen grains were mounted on glass
slides and identified and quantified under an Axiostar Carl Zeiss
microscope (400x and 1000x) using published pollen atlas Heusser (1971)
and the reference pollen collection of the Paleoecology Laboratory at
IPE-CSIC. Pollen counts include 300 terrestrial pollen grains excluding
paludal taxa. Phytoliths were obtained from twenty four samples of
approximately 2 gr of sediment. We followed the standard procedures for
phytoliths extraction (Piperno, 1988, 1994, 2006) using 36% HCl and 70%
HNO3 to remove carbonates and organic matter respectively. Phytoliths
were separated by heavy liquid flotation using ZnBr2 prepared to a
density of 2.3 g/cm3 (Dickau et al., 2013). Finally, phytoliths were
counted until obtaining a total of 200 per sample and they were
classified according to the International Code for Phytolith
Nomenclature (Madella et al. 2005). Diatoms were extracted
simultaneously with phytoliths. Digested aliquotes were mounted onto
permanent slides with Naphrax. Diatom valves were visualized under a
XXXXX microscope, with an oil immersion objective of 1000x. A minimum of
400 valves were counted for each level, with the exception of the base
levels X, Y, Z, where valves were scarce. Diatoms were identified to
species level with the LM images (1000x) taken using a digital SLR
camera (Canon EOS Rebel) attached to microscope AND SEM???. The
following literature was used for taxonomy and autoecology of the
encountered valves.

The TT3 chronology was reconstructed with thirteen AMS-radiocarbon dates
on charcoal and bulk sediment determined at DirectAMS laboratories, USA.
Results are presented in units of percent modern carbon (pMC) and the
uncalibrated radiocarbon age before present (relative to 1950 CE).
Radiocarbon ages were calibrated using the Southern Hemisphere
calibration curve (SHCal20, Hogg et al 2020). We used the package rbacon
v2.2 (Blaauw and Christen, 2011) for age--depth modeling and to
establish the deposition rates along the sequence. A change point
analysis was carried out to identify significant breaks in the time
series (Bai and Perron, 2003; Killick and Eckley, 2014). Stratigraphic
diagrams were analyzed and plotted using 'tidypaleo' v0.1.1, 'ggplot'
v3.3.5 and 'rioja' v0.9-26 packets in R 4.1.0 software.

# Results

## Sedimentary facie, geochemical and end-member mixing analysis 

## Phytoliths and Pollen analysis

## Diatom analysis

## Radiocarbon chronology

```{r fig.width=17, fig.height=11, echo=FALSE, dev="cairo_pdf",fig.align = "center",out.width = '100%'}

tar_read(strat_plot)

```



```{r echo = FALSE}

tar_read(chrono_table)

```



```{r echo = FALSE}

tar_read(chrono_plot)

```



# Discutions

# References {#references .unnumbered}



\clearpage

\begin{table}[!hbtp]
\caption{Sedimentological, compositional characteristics and depositional environment of LdM lacustrine facies.}
\resizebox*{0.98\textwidth}{!}{
\begin{tabular}{p{1cm}p{5.5cm}p{5.5cm}p{5.2cm}}
\hline\hline
\multicolumn{4}{l}{\textbf{Banded diatomaceous silts  (TOC $<$2$\%$; BioSi 18-19 $\%$)}}\\ [0.7ex] 
\hline
\textbf{Facies} & \textbf{Sedimentological properties} & \textbf{Composition and geochemistry} & \textbf{Depositional environment} \\ [0.7ex] 
\hline
\textbf{D1} & Dark to light brown, diatomaceous fine silt in cm- thick layers with gradational boundaries. & Low TOC (1.9 $\%$), TIC ($<$ 0.1 $\%$) and TS (0.24 $\%$) and moderately low BioSi (19.5 $\%$). High silicate-related elements (K, Ti, Rb, Sr and Zr); low Br/Ti, Ca/Ti, S/Ti and Fe/Mn.& Distal, profundal environments with intermediate clastic-volcanic input (ash fall processes and clastic run-off from the watershed after volcanic events).Higher clastic input in D2.\\ [0.5ex]
\hline
\textbf{D2} & Dark grey to brown, diatom-aceous, medium to coarse silt in cm-thick layers with gradational top and bottom boundaries. & Lower TOC (1.6 $\%$), TIC ($<$ 0.1), TS (0.22 $\%$) and BioSi (18.2 $\%$) than facies D1. High silicate-related elements (K, Ti, Rb, Sr and Zr); low Br/Ti, Ca/Ti, S/Ti and Fe/Mn.& Distal, profundal environments with higher clastic - volcanic input \\ [0.5ex]
\hline
\multicolumn{4}{l}{\textbf{Banded, diatomaceous, organic-rich silts (TOC $>$3$\%$; BioSi $>$ 20$\%$)}}\\ [0.7ex] 
\hline \hline
\textbf{Facies} & \textbf{Sedimentological properties} & \textbf{Composition and geochemistry} & \textbf{Depositional environment} \\ [0.7ex] 
\hline
\textbf{D3} & Banded, dm-thick layers of dark brown diatomaceous, organic rich medium to coarse silts with variable macrophyte content. & High TOC (3.8 $\%$) and BioSi ($21.7$ $\%$), moderate TS ($0.6$ $\%$) and low TIC ($< $$0.1$ $\%$). Relatively high Si, Fe and Br and low Ti, Rb, Sr, Zr, Ti, Br/Ti, Ca/Ti, Fe/Mn.& \multirow{3}{5cm}{Littoral to relatively deep environments dominated by organic deposition with low clastic input. Sequences (D3/D5,D4), (D5/D3/D4) and alternation D3/D4 represent transition from littoral, macrophyte-dominated (D3 and D5) to deeper environments (D4).}\\ [0.5ex]
\cline{1-3}
\textbf{D4} & Massive to banded, $<$ dm-thick layers of lighter brown diatom and organic rich fine silt. & High TOC ($3.2$$\%$) and BioSi ($21.4$ $\%$), moderate TS ($0.7$ $\%$) and low TIC ($<$ $0.1$ $\%$). Relatively high Si, Fe and Br and low Ti, Rb, Sr, Zr.&\\ [0.5ex]
\cline{1-3}
\textbf{D5} &Massive to banded, mm- to cm - thick layers of brown, macrophyte-rich coarse silt.& TS (1.4 $\%$) and presence of calcite (TIC = $0.1$$\%$). Higher TOC/TN. Moderate but fluctuating silicate-related elements (K, Ti, Rb, Sr and Zr), high Br/Ti, Ca/Ti, and S/Ti. High Mn values, and low Fe/Mn ratios.&\\ [0.5ex]
\hline
\multicolumn{4}{l}{\textbf{Laminated diatomaceous organic-rich silts(TOC 2-3.5$\%$; BioSi ca. 20 $\%$)}}\\ [0.7ex] 
\hline \hline
\textbf{Facies} & \textbf{Sedimentological properties} & \textbf{Composition and geochemistry} & \textbf{Depositional environment} \\ [0.7ex] 
\hline
\textbf{D6} &Laminated, dark brown, organic-rich silt composed of three main laminae type: a) greenish grey, diatomaceous medium to fine silt; b) dark brown, organic-rich fine silt with diffuse upper and lower boundaries and c) brown, macrophyte–rich, coarse silt, with sharp, erosional lower boundaries. Facies D6b contains more frequent endogenic calcite crystals.& High TOC (3.1$\%$), BioSi (19.7$\%$) and TS (1.4$\%$, peaks up to 6$\%$ in D6c) and common presence of calcite (TIC $>=$ 0.1). Moderate but fluctuating silicate-related elements (K, Ti, Rb, Sr and Zr); high Br/Ti, Ca/Ti, and S/Ti. High Mn values and low Fe/Mn ratios.& Distal, frequently oxic environments dominated by organic deposition and higher clastic input. Some periods with endogenic calcite formation\\ [0.5ex]
\hline
\end{tabular}}
\end{table}
